---
title: "football"
output: pdf_document
date: "2023-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Import Libraries
library(dplyr)
library(tidyr)
library(tidyverse)
```
```{r}
# Code
data_df = read_csv('data/cleaned_match_data.csv')

# Removed date because it doesnt add any value to data
data_df = data_df[-c(1)]

# Convert Datatype
data_df$HTR = as.factor(data_df$HTR)
data_df$FTR = as.factor(data_df$FTR)
data_df$HomeTeam = as.factor(data_df$HomeTeam)
data_df$AwayTeam = as.factor(data_df$AwayTeam)
str(data_df)
```

```{r}
numeric_col = unlist(lapply(data_df, is.numeric))

cat("Numerical Columns within dataset: ", colnames(data_df[, numeric_col]),"\n")

cat("Categorical Columns within dataset: ", colnames(data_df[, !numeric_col]))

data_df[,numeric_col] = sapply(data_df[,numeric_col],as.integer)
```

```{r fig.height=6, fig.width=10, warning=FALSE}
# Distributions Analysis
gg <- 
  ggplot(gather(data_df[,!numeric_col], cols, value), aes(x = value)) + 
       geom_histogram(stat="count") + facet_wrap(.~cols, scales="free", nrow=2)

ggsave(
  file.path("images", "cat_distribution.png"),
  plot = gg,
  width = 12, 
  height = 8, 
  units = "in",
  dpi = 200
)

gg

gg <- 
  ggplot(gather(data_df[, numeric_col], cols, value), aes(x = value)) + 
  geom_histogram() +
  facet_wrap(.~cols, scales="free", nrow=5)

ggsave(
  file.path("images", "num_distribution.png"),
  plot = gg,
  width = 12, 
  height = 8, 
  units = "in",
  dpi = 200
)

gg
```
```{r fig.height=14, fig.width=14, warning=FALSE}
# Correlation Plots
upper_panel<-function(x, y){
    points(x,y, pch=19, col=c("darkblue", "darkorange", "violet")[data_df$FTR])
}

lower_panel <- function(x, y){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y), digits=2)
    txt <- paste0("R = ", r)
    text(0.5, 0.5, txt)
}

num_data = data_df[,numeric_col]
num_data = cbind(num_data, FTR=data_df$FTR)

# pairs(num_hdp_data[c(1:6)],
#       upper.panel = upper_panel,
#       lower.panel = lower_panel
#       )
# pairs(num_hdp_data[c(7:12)],
#       upper.panel = upper_panel,
#       lower.panel = lower_panel
#       )
# pairs(num_hdp_data[c(13:18)],
#       upper.panel = upper_panel,
#       lower.panel = lower_panel
#       )
# pairs(num_hdp_data[c(19:21)],
#       upper.panel = upper_panel,
#       lower.panel = lower_panel
#       )

gg <- pairs(num_data,
      upper.panel = upper_panel,
      lower.panel = lower_panel)

ggsave(
  file.path("images", "correlation.png"),
  plot = gg,
  width = 14, 
  height = 14, 
  units = "in",
  dpi = 200
)

gg
```

```{r}
total_team_matches <- 

home_team_wins <- as.data.frame(table(filter(data_df, FTR=="H")$HomeTeam))
away_team_wins <- as.data.frame(table(filter(data_df, FTR=="A")$AwayTeam))
draw_team_wins <- filter(data_df, FTR=="D")[c("HomeTeam", "AwayTeam")]

library(reshape2)
draw_team_wins <- melt(draw_team_wins, id.var=c("HomeTeam", "AwayTeam"), variable.name = "Total Draws")

team_wins <- cbind(home_team_wins,away_team_wins$Freq)
colnames(team_wins) <- c("TeamName", "HomeWins", "AwayWins")
```

```{r fig.height=14, fig.width=14}
team_wins_plot <- melt(team_wins, id.vars = 1)

ggplot(aes(x=variable, y=value), data=team_wins_plot)+
  geom_bar(aes(fill=variable), stat="identity", position="dodge")+
  facet_wrap(vars(team_wins$TeamName))

barplot(team_wins, legend = c("HomeWins", "AwayWins"), beside=TRUE)
```

```{r}
install.packages('ggplot2')

library(ggplot2)
install.packages('reshape2')

library(reshape2)
# heat map correlation
title = "correlation heat map" 
heat_map = qplot(x=Var1, y=Var2, data=melt(cor(df_subset, use="p")), fill=value, geom="tile") +  
      scale_fill_gradient2(limits=c(-1, 1))
heat_map <- heat_map + theme(axis.title.x=element_blank(), axis.text.x=element_blank()  
                     , axis.ticks=element_blank())
heat_map <- heat_map + ggtitle(title)  
heat_map  
```
```{r}
install.packages('gclus')
library(gclus)
#scatter plot matrix

dta.col <- dmat.color(correlation_matrix) # get colors
# reorder variables so those with highest correlation
# are closest to the diagonal
cpairs(df_subset, ord, panel.colors=dta.col, gap=.5,
main="Variables Ordered and Colored by Correlation" )
```

```{r}
install.packages('ggplot')
library(ggplot2)

df_subset2 <- df[ , c(4,5,6,7,8,9,10:21)] 

df_subset2$FTR2 <- factor(df_subset2$FTR,levels=c("0", "1","2"),
labels = c("A", "D","H"))

df_subset2$HTR2 <- factor(df_subset2$HTR,levels=c("A", "D","H"),
labels = c("0", "1","2"))

ggplot(df_subset2, aes(x=df_subset2$FTHG,color=df_subset2$FTR2)) + geom_histogram(POSITION='dodge',binwidth=5)
```

```{r}
table(df_subset2$FTAG)
```
```{r}
table(df_subset2$FTHG)

```
```{r}

# Create a histogram:
hist(df_subset2$FTAG, breaks = 5, main = "Full Time Away Team Goals", 
     xlab = "No. of Goals", ylab = "Frequency")

```

```{r}
# Create a histogram of the data with 5 bins
hist(df_subset2$FTHG, breaks = 5, main = "Full Time Home Team Goals", 
     xlab = "No. of Goals", ylab = "Frequency")
```

```{r}
# Load the ggplot2 package
library(ggplot2)

# Create two sample vectors of data
data1 <- df_subset2$FTAG
data2 <- df_subset2$FTHG

# Combine the data into a data frame
df <- data.frame(Data = c(data1, data2), Group = c(rep("FTAG", length(data1)), 
                                                   rep("FTHG", length(data2))))

# Create a histogram of the data with separate colors for each group
ggplot(df, aes(Data, fill = Group)) + 
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  ggtitle("Histogram of Full Time Goals") +
  xlab("No. of Goals in a game") + ylab("Frequency")

```

This data is for full time goals. Blue represents the full time home team goals.
Seems like home team is outperforming the away team in all the matches.

```{r}
# Load the ggplot2 package
library(ggplot2)

# Create two sample vectors of data
data1 <- df_subset2$HTAG
data2 <- df_subset2$HTHG

# Combine the data into a data frame
df <- data.frame(Data = c(data1, data2), Group = c(rep("FTAG", length(data1)), 
                                                   rep("FTHG", length(data2))))

# Create a histogram of the data with separate colors for each group
ggplot(df, aes(Data, fill = Group)) + 
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  scale_fill_manual(values = c('black', 'red')) +
  ggtitle("Histogram of Full Time Goals") +
  xlab("No. of Goals in a game") + ylab("Frequency")
```
Even in case of half time games, the home team is outperforming the away team.

```{r}
# Load the ggplot2 package
library(ggplot2)

# Create two sample vectors of data
data1 <- df_subset2$HF
data2 <- df_subset2$AF

# Combine the data into a data frame
df <- data.frame(Data = c(data1, data2), Group = c(rep("HF", length(data1)), rep("AF", length(data2))))

# Create a histogram of the data with separate colors for each group
ggplot(df, aes(Data, fill = Group)) + 
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  scale_fill_manual(values = c(rgb(0, 114, 178, maxColorValue = 255), 
                                rgb(230, 159, 0, maxColorValue = 255))) +
  ggtitle("Histogram of Fouls") +
  xlab("No. of Fouls in a game") + ylab("Frequency")
```
Seems like, home team is also having more fouls than the away team.

```{r}
# Load the ggplot2 package
library(ggplot2)

# Create two sample vectors of data
data1 <- df_subset2$HY
data2 <- df_subset2$AY

# Combine the data into a data frame
df <- data.frame(Data = c(data1, data2), Group = c(rep("HY", length(data1)), 
                                                   rep("AY", length(data2))))


# Create a histogram of the data with separate colors for each group
ggplot(df, aes(Data, fill = Group)) + 
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  scale_fill_manual(values = c(rgb(0, 114, 178, maxColorValue = 255), 
                                rgb(230, 159, 0, maxColorValue = 255))) +
  ggtitle("Histogram of Yellow Cards") +
  xlab("No. of Yellow cards in a game") + ylab("Frequency")
```

```{r}
# Load the ggplot2 package
library(ggplot2)

# Create two sample vectors of data
data1 <- df_subset2$HR
data2 <- df_subset2$AR

# Combine the data into a data frame
df <- data.frame(Data = c(data1, data2), Group = c(rep("HR", length(data1)), rep("AR", length(data2))))

# Create a histogram of the data with separate colors for each group
ggplot(df, aes(Data, fill = Group)) + 
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  scale_fill_manual(values = c(rgb(0, 114, 178, maxColorValue = 255), 
                                rgb(230, 159, 0, maxColorValue = 255))) +
  ggtitle("Histogram of Red cards") +
  xlab("No. of Red cards in a game") + ylab("Frequency")
```
Away team gets more red cards.

```{r}
# Fouls Vs FTR (0=away wins, 1=draw, 2=home wins)
p <- ggplot(df_subset2, aes(HF, AF),colour = factor(FTR)) + geom_point()
# With one variable

p + facet_grid(. ~ FTR)
```

```{r}
# Create the plot
p <- ggplot(df_subset2, aes(FTHG, FTAG)) + geom_point(aes(color = factor(FTR)))

# Add facet grid with colored backgrounds
p + facet_grid(. ~ FTR, scales = "free", space = "free", switch = "x") +
  scale_color_manual(values = c("red", "blue", "green")) +
  theme(strip.background = element_rect(fill = NA, colour = NA), 
        strip.text = element_text(colour = "black", size = 10, face = "bold"))

```
0 = Away team wins
1 = draw
2 = home team wins
```{r}
# Create a mosaic plot
mosaicplot(table(df_subset2$FTR,df_subset2$FTAG ), color = c("black", "gray"))
```
When away team wins (0), the maximum goals are = 2.
When home team wins (2), maximum goals are = 1!!!

```{r}
table(df$HomeTeam)

```


```{r}
# Get count of each category
count_table <- table(df$AwayTeam)

# Sort the table in ascending order by count
sorted_table <- count_table[order(count_table)]

# Print the sorted table
print(sorted_table)

```

For Away teams, we have maximum data for Tottenham, Man City, Everton, Chelsea, 
Arsenal. 
We have least data for Nott'm Forest.

```{r}
# Create a sample data frame
df1 <- data.frame(team = df$HomeTeam,
                 result = df$FTR)

# Calculate the number of wins and losses for each team
library(dplyr)

summary <- df1 %>%
  group_by(team) %>%
  summarize(wins = sum(result == "H"), losses = sum(result == "A"),
            draw = sum(result=='D'),win_percent = round(wins / n() * 100, 2),
            loss_percent = round(losses / n() * 100, 2),
            draw_percent = round(draw / n() * 100, 2)) %>%
  arrange(desc(win_percent))
# Print the summary
print(summary)
```

Highest winning percentage is for Man City in home teams.

```{r}
# Create a sample data frame
df1 <- data.frame(team = df$AwayTeam,
                 result = df$FTR)

# Calculate the number of wins and losses for each team
library(dplyr)

summary <- df1 %>%
  group_by(team) %>%
  summarize(wins = sum(result == "A"), losses = sum(result == "H"),
            draw = sum(result=='D'),win_percent = round(wins / n() * 100, 2),
            loss_percent = round(losses / n() * 100, 2),
            draw_percent = round(draw / n() * 100, 2)) %>%
  arrange(desc(win_percent))
# Print the summary
print(summary)
```

In Away Teams too Man CIty is at the top. However, Arsenal which was at second 
position has now dropped to fifth position.
```{r}
# Create a sample data frame
df1 <- data.frame(team = df$HomeTeam,
                 result = df$HS)

# Calculate the number of wins and losses for each team
library(dplyr)

summary <- df1 %>%
  group_by(team) %>%
  summarize(A = sum(result >0 & result<=10), B = sum(result >10 & result<=20),
            C = sum(result >20 ))%>%
  arrange(desc(C))
# Print the summary
print(summary)
```
This represents that Man CIty had the most home team shots.
As we saw before in correlation plots, home team shots are correlated to a home
team winning. 

```{r}
# Create a sample data frame
df1 <- data.frame(team = df$HomeTeam,
                 shots = df$HS, results = df$FTR)

# Calculate the number of wins and losses for each team
library(dplyr)

summary <- df1 %>%
  group_by(team) %>%
  summarize(A = sum(shots >0 & shots<=10), B = sum(shots >10 & shots<=20),
            C = sum(shots >20 ), wins = sum(results == "H"), 
            losses = sum(results == "A"),
            draw = sum(results == 'D'))%>%
  arrange(desc(C))
# Print the summary
print(summary)

```

